SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2014-05-05 paper/paper.tex
paper: updates.

In this section, we motivate in more detail the challenges in shrinking
counterexamples by comparing manual approaches using QuickCheck to SmartCheck.
(We focus on shrinking rather than generalization here since counterexample
generalization is unique to SmartCheck.)  We will show how a small data type and
simple property can result in large counterexamples without any shrinking.  Then
we show the difficulty in designing an efficient shrink implementation manually,
which is required when using QuickCheck; we will walk through a couple of poor
designs before arriving at a ``canonical'' manual solution.  However, even the
canonical solution produces significantly larger counterexamples and takes
significantly longer to produce them than SmartCheck, even though SmartCheck's
shrinking strategy is completely automatic and generic.

Consider the example in Figure~\ref{fig:initial}.\footnote{All examples and
  algorithms in this paper are presented in Haskell 2010~\cite{haskell2010} plus
  the language extensions of existential type quantification and scoped type
  variables.}  Data type \ttp{T} is a product type containing five lists of
signed 8-bit integers.  For our purposes, it is not important what \ttp{T}
models, but for concreteness, suppose it models the values in five linked lists.

@@

Despite the simplicity of the example, a typical counterexample returned by
QuickCheck can be large.  With standard settings and no shrinking, the median
counterexample discovered contains just under 70 \ttp{Int16} values, and
counterexamples can contain over 100 values.  Thus, it pays to define
\ttp{shrink}!

@@

The first way we will control the complexity is focusing on structurally
shrinking the data type rather than shrinking the base types.  We prevent
QuickCheck from using its default shrink instance for \ttp{Int16} by defining a
newtype:
%
\begin{code}
newtype T = T { getInt :: Int16 }
\end{code}
%
\noindent
Without shrinking the \ttp{Int16}s, the performance is dramatically improved.

Next, a programmer might try to control the complexity by truncating lists using
truncation where
%
\begin{code}
take n ls
\end{code}
%
returns the first $n$ elements \ttp{ls}.  The trade-off is quicker shrinking
with a lower probability of finding a smaller counterexample.  For example, we
might redefine shrink as follows:
%
\begin{samepage}
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- tk w
                , b <- tk i0, c <- tk i1
                , d <- tk i2, e <- tk i3 ]
  where tk x = take 10 (shrink x)
\end{code}
\end{samepage}
%
\noindent
Call this version ``truncated shrink''.  Truncation controls the blowup of the
input-space, but the downside is that potentially smaller counterexamples may be
omitted.

Someone who understands the semantics of QuickCheck's \ttp{shrink}
implementation defines a \ttp{shrink} instance as follows:\footnote{The
  following approach was suggested to the author by John Hughes.}
%
\begin{code}
shrink (T w i0 i1 i2 i3) = map go xs
  where xs = shrink (w, i0, i1, i2, i3)
        go (w', i0', i1', i2', i3')
           = T w' i0' i1' i2' i3'
\end{code}
%
\noindent
This ``tuple shrink'' definition does not suffer the same shortcomings: it
shrinks a value even if it contains an empty list, and the combinatorial blowup
of shrink candidates is avoided, since a pair \ttp{(a,b)} is shrunk by
attempting to shrink \ttp{a} while holding \ttp{b} constant, then attempting to
shrink \ttp{b} while holding \ttp{a} constant (and is generalized from pairs to
arbitrary tuples).  Thus, each argument to \ttp{T} is independently shrunk.
Still, using this definition, from some initial counterexamples, shrinking may
result in a final counterexample containing just over 100 \ttp{Int16} values.

SmallCheck is another testing framework for Haskell for which shrinking is
irrelevant: SmallCheck is guaranteed to return a smallest counterexample, if
one exists~\cite{sc}.  SmallCheck does this by enumerating all possible inputs,
ordered from smallest to largest, up to some user-defined bound.  While
SmallCheck is effective for testing many programs and properties (in accordance
with the \emph{small scope hypothesis}~\cite{jackson}), counterexamples to even
relatively simple properties may be infeasible to discover due to input-space
explosion.

%% \begin{figure}
%%   \begin{code}
%% instance Serial Int16 where
%%   series d = drawnFrom [(-d')..d']
%%     where d' = fromIntegral d

%% instance Serial A where series = cons1 A
%% instance Serial B where series = cons4 B
%%   \end{code}
%%   \caption{SmallCheck instances for the data types in
%%     Figure~\ref{fig:initial}.}
%%   \label{fig:smallshrink}
%% \end{figure}

For a product type, SmallCheck must check values produced by taking the
cross-product of the type's fields at a given depth.  Unfortunately, using
default instances, SmallCheck fails long before it reaches a depth required to
discover a counterexample.  (A related library named Feat combines some aspects
of SmallCheck and QuickCheck~\cite{feat}; we discuss it in
Section~\ref{sec:related}).

@@

\begin{table}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline
           & QC none     & QC trunc     & QC tuple    & SmartCheck\\
\hline \hline
Mean       & 81, 0.013s  & 40, 0.231s   & 13, 0.228s  & 6, 0.022s\\
\hline
Std. dev.  & 21, 0.001   & 21, 1.742    & 7,  0.962   & 4, 0.001s\\
% Median     & 0.009s, 67  & 0.17s, 33  & 1.81s, 13  & 0.10s, 6\\
%% \hline
%% MAD        & 0.02s, 13  & 0.09s, 8   & 0.62s, 3   & 0.02s, 2\\
\hline
95\%       & 116, 0.025s  & 84, 0.283s  & 26, 0.501s  & 13, 0.039s\\
\hline
    \end{tabular}
  \end{center}
  \caption{Summarizing data for the graph in Figure~\ref{fig:graphs}. Entries
    contain execution time (in seconds) and counterexample sizes (counting
    \ttp{Int16} values).}
  \label{table:results}
\end{table}

To motivate the benefits of SmartCheck reduction algorithm, consider
Table~\ref{table:results} and the corresponding graphs in
Figure~\ref{fig:graphs}, giving the results for using QuickCheck and SmartCheck
on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck, since as
noted, it is not even feasible for the example with default
instances).\footnote{All results reported in the paper are with a
  GHC-7.6.3-compiled program, using \ttp{-O2}, on a 2.8 GHz 4 core i7 with 16GB
  memory under OSX v. 10.9.2.}  We graph the final counterexample size and
execution time.

The experiments show three approaches to shrinking with QuickCheck: no shrinking
(QC none), to provide a baseline, truncated shrinking (QC trunc), as described
above, in which each list is truncated to 10 elements and \ttp{Int16} values are
not shrunk, and tuple shrinking (QC tuple), also described above, and
\ttp{Int16} values are also not shrunk.

With each shrinking strategy, we generate 1000 runs in which a random
counterexample is discovered (by QuickCheck), and then shrunk.  We plot the
number of runs resulting in a shrunk counterexample of a given size.

@@

The execution time includes the time to
discover an initial counterexample as well as time to shrink it.

SmartCheck produces smaller counterexamples than the other approaches, with a
very small tail.

More dramatic, however, are the differences in execution time.  The execution
times for SmartCheck include the initial counterexample generation time required
by QuickCheck (QC none); thus the time required by SmartCheck to shrink a
counterexample is on average just two hundredths of a second (0.10 - 0.08).  As
can be seen in the graphs, SmartCheck's execution time scales linearly (like
QuickCheck without shrinking).  Truncation and tuple shrinking produce very long
tails; the longest execution time recorded for truncation shrinking and tuple
shrinking are 88 seconds and 110 seconds, respectively, although these outliers
can vary wildly.  SmartCheck is largely insensitive to the size of the original
counterexample provided.

@@

Finally, these performance results for SmartCheck are fairly generalizable to
other types and properties, since the algorithm is not dependent on the specific
program or property being checked.  However, the results are particularly
pronounced for counterexamples that are particularly ``deep'' or ``broad'' in
terms of their tree structure.  %% For programs and properties for which small
%% counterexamples are easily found, QuickCheck's shrinking is perfectly
%% acceptable.  Furthermore, user-defined shrink methods may vary significantly
%% based on the property and counterexample type, so it is not always clear which
%% implementation to compare against.
We now describe in detail SmartCheck's algorithm for generic shrinking.

@@

\subsection{Benchmarks}

No set of testing benchmarks exists over which to compare different test-case
generation and minimization approaches.  Therefore, we have collected a small
number of benchmarks, in addition to the more involved case-studies described
earlier in this section.

The benchmarks presented, in addition to the motivating example presented in
Section~\ref{sec:example}, are

@@

\noindent
All benchmarks can be found at
\url{https://github.com/leepike/SmartCheck/tree/master/regression}.  Benchmarks
are against
QuickCheck~2.7.3.\footnote{\url{http://hackage.haskell.org/package/QuickCheck-2.7.3}}
The benchmarks only compare the size of shrunk values and the time required to
generate them using standard arguments to QuickCheck and SmartCheck.  We do not
benchmark the generalization algorithms, since there is no analog in QuickCheck
or other testing frameworks.

\todo{include std dev.}
\todo{fill in benchmarks!}

@@

\todo{note that 95 is smaller than average for heap}

The \emph{Reverse} benchmark essentially provides a lower-bound on the benefit
of using SmartCheck to discover small counterexamples, since a list of length
two falsifies the property (SmartCheck is still useful in generalizing the
counterexample, however!).  Perhaps surprisingly, SmartCheck still slightly
outperforms QuickCheck.  The other two benchmarks are slightly larger examples,
where SmartCheck's benefits are more pronounced.  But again, we emphasize the
most pronounced benefit comes in much larger examples, like described in
Section~\ref{sec:nlp}.

@@

@@

Feat finds small counterexamples effectively, but a limitation compared to
QuickCheck is that value generation is applicative rather than monadic.
Consequently, defining custom generators (i.e., instances for \ttp{arbitrary})
is constrained.  On the other hand, generators can be derived
automatically with Feat.

We have benchmarked Feat as well on the example in the introduction
(Figure~\ref{fig:initial}).  It performs quite competitively to SmartCheck
in terms the size of counterexamples discovered: the mean and median number of
\ttp{N} constructors in the counterexamples returned by Feat are both seven. \todo{check}
However, Feat performs less well with respect to how long counterexample
discovery takes: the mean time taken by Feat is 5.60 seconds, with a long tail:
five percent of runs take over 14 seconds (the mean time taken by
SmartCheck is 0.12 seconds, and 99\% of runs take less than 0.30 seconds).

@@

\todo{mention memoization (generalized tries)}

\todo{tail-recursive algorithm is fast and easy to port to imperative languages}

>>>>

In this section, we motivate in more detail the challenges in shrinking
counterexamples by comparing manual approaches using QuickCheck to SmartCheck.
(We focus on shrinking rather than generalization here since counterexample
generalization is unique to SmartCheck.)  We will show how a small data type and
simple property can result in large counterexamples without any shrinking.  Then
we show the difficulty in designing an efficient shrink implementation manually,
which is required when using QuickCheck; we will show a poor
design before arriving at a ``canonical'' manual solution.  However, even the
canonical solution produces significantly larger counterexamples and takes
significantly longer to produce them than SmartCheck, even though SmartCheck's
shrinking strategy is completely automatic and generic.

Consider the example in Figure~\ref{fig:initial}.\footnote{All examples and
  algorithms in this paper are presented in Haskell 2010~\cite{haskell2010} plus
  the language extensions of existential type quantification and scoped type
  variables.}  Data type \ttp{T} is a product type containing five lists of
signed 8-bit integers.  %% For our purposes, it is not important what \ttp{T}
%% models, but for concreteness, suppose it models the values in five linked lists.

@@

Despite the simplicity of the example, a typical counterexample returned by
QuickCheck can be large.  With standard settings and no shrinking, the average
counterexample discovered contains just over 80 \ttp{Int16} values, and over
five percent contain over 100 values.  Thus, it pays to define
\ttp{shrink}!

@@

The first way we will control the complexity is focusing on structurally
shrinking the data type rather than shrinking the base types.  We prevent
QuickCheck from using its default shrink instance for \ttp{Int16} by defining a
newtype:
%
\begin{code}
newtype J = J { getInt :: Int16 }
\end{code}
%
\noindent
Next, a programmer might try to control the complexity by truncating lists using
truncation where
%
\begin{code}
take n ls
\end{code}
%
returns the first $n$ elements \ttp{ls}.  The trade-off is quicker shrinking
with a lower probability of finding a smaller counterexample.  For example, we
might redefine shrink as follows:
%
\begin{samepage}
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- tk w
                , b <- tk i0, c <- tk i1
                , d <- tk i2, e <- tk i3 ]
  where tk x = take 10 (shrink x)
\end{code}
\end{samepage}
%
\noindent
Call this version ``truncated shrink''.  Truncation controls the blowup of the
input-space, but the downside is that potentially smaller counterexamples may be
omitted.

Someone who understands the semantics of QuickCheck's \ttp{shrink}
implementation defines a \ttp{shrink} instance as follows:\footnote{The
  following approach was suggested to the author by John Hughes.}
%
\begin{code}
shrink (T w i0 i1 i2 i3) = map go xs
  where xs = shrink (w, i0, i1, i2, i3)
        go (w', i0', i1', i2', i3')
           = T w' i0' i1' i2' i3'
\end{code}
%
\noindent
This ``tuple shrink'' definition does not suffer the same shortcomings: it
shrinks a value even if it contains an empty list, and the combinatorial blowup
of shrink candidates is avoided, since a pair \ttp{(a,b)} is shrunk by
attempting to shrink \ttp{a} while holding \ttp{b} constant, then attempting to
shrink \ttp{b} while holding \ttp{a} constant (and is generalized from pairs to
arbitrary tuples).  Thus, each argument to \ttp{T} is independently shrunk.

Finally, with this combination of shrinking tuples and disabling shrinking for
\ttp{Int16} values, we have an efficient and effective shrinking approach.

@@

\begin{table}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline
           & QC none     & QC trunc     & QC tuple    & SmartCheck\\
\hline \hline
Mean       & 83, 0.013s  & 40, 0.064s   & 10, 0.031s  & 6, 0.021s\\
\hline
Std. dev.  & 22, 0.014   & 20, 0.334    & 4,  0.039   & 4, 0.009\\
% Median     & 0.009s, 67  & 0.17s, 33  & 1.81s, 13  & 0.10s, 6\\
%% \hline
%% MAD        & 0.02s, 13  & 0.09s, 8   & 0.62s, 3   & 0.02s, 2\\
\hline
95\%       & 119, 0.023s  & 84, 0.153s  & 16, 0.052s  & 13, 0.036s\\
\hline
    \end{tabular}
  \end{center}
  \caption{Summarizing data for the graph in Figure~\ref{fig:graphs}. Entries
    contain execution time (in seconds) and counterexample sizes (counting
    \ttp{Int16} values).}
  \label{table:results}
\end{table}

The results of our discussion above are summarized in Table~\ref{table:results}
and the corresponding graph in Figure~\ref{fig:graphs}. They show three
approaches to shrinking with QuickCheck: no shrinking (QC none), to provide a
baseline, truncated shrinking (QC trunc) with \ttp{Int16} shrinking disabled, in
which each list is truncated to 10 elements, and tuple shrinking (QC tuple),
also with \ttp{Int16} shrinking disabled.  We omit SmallCheck, since as noted,
it is not even feasible for the example with default instances.\footnote{All
  results reported in the paper are with a GHC-7.6.3-compiled program, using
  \ttp{-O2}, on a 2.8 GHz 4~core i7 with 16GB memory under OSX v. 10.9.2.}  We
graph the final counterexample size, measured in the total number of \ttp{Int16}
values, and execution time for counterexample discovery and shrinking.
Figure~\ref{fig:graphs} shows how many of the 1000 runs resulted in a final
value of a particular size, fitting the data with a Bezier curve.

@@

SmartCheck produces consistently small counterexamples and does so
quickly---even smaller and faster than the best hand-tuned solution---all
without writing any problem-specific code.  Note that the ``no shrinking''
example provides a lower-bound on execution time (only including counterexample
discovery time) and an upper-bound on counterexample size.  In this example, the
hand-written ``tuple shrink'' algorithm is only slightly slower and produces
slightly larger results.  In Section~\ref{sec:benchmarks}, we provide additional
benchmarks, but now we turn to the design of SmartCheck.

%% More dramatic, however, are the differences in execution time.  The execution
%% times for SmartCheck include the initial counterexample generation time
%% required by QuickCheck (QC none); thus the time required by SmartCheck to
%% shrink a counterexample is on average just two hundredths of a second (0.10 -
%% 0.08).  As can be seen in the graphs, SmartCheck's execution time scales
%% linearly (like QuickCheck without shrinking).  Truncation and tuple shrinking
%% produce very long tails; the longest execution time recorded for truncation
%% shrinking and tuple shrinking are 88 seconds and 110 seconds, respectively,
%% although these outliers can vary wildly.  SmartCheck is largely insensitive
%% to the size of the original counterexample provided.

@@

%% Finally, these performance results for SmartCheck are fairly generalizable to
%% other types and properties, since the algorithm is not dependent on the
%% specific program or property being checked.  However, the results are
%% particularly pronounced for counterexamples that are particularly ``deep'' or
%% ``broad'' in terms of their tree structure.  %% For programs and properties
%% for which small counterexamples are easily found, QuickCheck's shrinking is
%% perfectly acceptable.  Furthermore, user-defined shrink methods may vary
%% significantly based on the property and counterexample type, so it is not
%% always clear which implementation to compare against.
We now describe in detail SmartCheck's algorithm for generic shrinking.

@@

\subsection{Benchmarks}\label{sec:benchmarks}

Unfortunately, no set of testing benchmarks exists over which to compare
different test-case generation and minimization approaches.  Therefore, we have
collected a small number of benchmarks, in addition to the more involved
case-studies described earlier in this section.

The benchmarks presented, in addition to the motivating example presented in
Section~\ref{sec:example}, compare standard SmartCheck against hand-written
shrink implementations.  The benchmarks are as follows:

@@

\noindent
All benchmarks can be found
online.\footnote{\url{https://github.com/leepike/SmartCheck/tree/master/regression}}
We compare the size of the final counterexample returned (by counting
constructors) and the time required for counterexample generation and shrinking
in seconds.  The results are presented in Table~\ref{table:results}.  Again, we
summarize the mean, median, standard deviation, and the results at the 95th
percentile.  (While we provide the standard deviations, note that the plots are
not necessarily Gaussian.)

@@

The \emph{Reverse} benchmark essentially provides a lower-bound on the benefit
of using SmartCheck to discover small counterexamples, since a list of length
two falsifies the property (SmartCheck is still useful in generalizing the
counterexample, however!).  Perhaps surprisingly, SmartCheck still slightly
outperforms QuickCheck.  The other two benchmarks have much larger counterexamples,
where SmartCheck's benefits are more pronounced.  For example, consider the
parser example in which the initial counterexample to shrink may be in a large
parse tree.  SmartCheck performs significantly better than optimal hand-written
shrink instances, but even SmartCheck can sometimes return large
counterexamples---this is why the mean counterexample size is larger than the
size at the 95th percentile: there are a small number of large values not
successfully shrunk by SmartCheck.

@@

SmallCheck is another testing framework for Haskell for which shrinking is
irrelevant: SmallCheck is guaranteed to return a smallest counterexample, if
one exists~\cite{sc}.  SmallCheck does this by enumerating all possible inputs,
ordered from smallest to largest, up to some user-defined bound.

While SmallCheck is effective for testing many programs and properties (in
accordance with the \emph{small scope hypothesis}~\cite{jackson}),
counterexamples to even relatively simple properties may be practically
infeasible to discover due to the size of the input space.  For example,
SmallCheck does not find a counterexample to the example presented in
Section~\ref{sec:example} after running it for several minutes.

@@

Feat is used to discover counterexamples, not shrink them.  However, shrinking
is less necessary with Feat, since discovered counterexamples are often small,
if one is found.  For example, on the overflow example in
Section~\ref{sec:example}, with a limit of 100 test cases, Feat finds a
counterexample just two percent of the time, whereas QuickCheck finds one nearly
100\%.  Even at a limit of 10000, a counterexample is found about 50\% of the
time (with a correspondingly longer search time).  Sampling from a uniform
distribution does not work so well here.  Feat does a better job of discovering
counterexamples in the parser benchmark, but the size of the average
counterexample contains 500 constructors, with a standard deviation of 500
(compared with 16 and 75, respectively, for SmartCheck).  Still, Feat is
powerful at what it does well and can be seamlessly used with SmartCheck, since
it just defines the \ttp{arbitrary} method.

%% When Feat finds a counterexample, it is a small one, rendering shrinking
%%   A limitation compared to QuickCheck (and SmartCheck, since
%% SmartCheck uses QuickCheck as a front-end for initial counterexample discovery)
%% is that value generation is applicative rather than monadic.  Consequently,
%% defining custom generators (i.e., instances for \ttp{arbitrary}) is constrained.
%% On the other hand, generators can be derived automatically with Feat.  The main
%% limitation, however

%% We have benchmarked Feat as well on the example in the introduction
%% (Figure~\ref{fig:initial}).  It performs quite competitively to SmartCheck in
%% terms the size of counterexamples discovered: the mean and median number of
%% \ttp{N} constructors in the counterexamples returned by Feat are both
%% seven. \todo{check} However, Feat performs less well with respect to how long
%% counterexample discovery takes: the mean time taken by Feat is 5.60 seconds,
%% with a long tail: five percent of runs take over 14 seconds (the mean time taken
%% by SmartCheck is 0.12 seconds, and 99\% of runs take less than 0.30 seconds).

@@

%% \todo{mention memoization (generalized tries)}

%% \todo{tail-recursive algorithm is fast and easy to port to imperative languages}
